Deciding Bitvector Termination with Program Synthesis

### 2014-07-06 papers/termination/paper.tex

All we need to do in order to use {\bf [CT]} in any context %environment and regardless of the complexity of the loops's body 
is generate constraints over-approximating %abstracting 
the environment and the loop's body. %: for termination, we need over-approximating abstractions, whereas for non-termination 
%we need under-approximations. 
%
As the generation of such constraints is standard and covered by several
other
works~\cite{DBLP:conf/pldi/GrebenshchikovLPR12,DBLP:conf/pldi/GulwaniSV08},
we will not provide the full algorithm, but rather illustrate it through the
example in Figure~\ref{fig:environment-model}.  Let us assume that we have
already proven the termination of loops $L_1$ and $L_3$, and we are now
trying to prove that $L_2$ terminates, which we encode as the conjunction of
constraints on the right-hand side of the figure:
%
\begin{itemize}
\item $W_1$, $W_2$ and $W_3$ are (over-approximating) inductive invariants for $L_1$, $L_2$ and $L_3$, respectively.
\item $S$ is a summary of $L_3$ that over-approximates the transitive closure of its transition relation.
\item If the formula is satisfiable, then $R$ is a ranking function for $L_2$.
 \end{itemize}

%the lines marked as $*$ over-approximate the states in which loop 2 can start execute.
%% contains a non-trivial program
%% along with the constraint system
%% encoding the termination of the outer loop.  Since the program contains sequential composition
%% of loops and nested loops, the constraint
%% system involves termination invariants $W_1, W_2, W_3$ and the inner loop summary $S$.

Note that despite its intricacy, the constraint system is still a formula in the synthesis fragment.
Since the outer loop of this program does indeed terminate, the system is satisfiable, as witnessed by:
\begin{align*}
W_1(\langle v, x, y, z \rangle ) & \triangleq x \leq y + 1 \\
W_2(\langle v, x, y, z \rangle ) & \triangleq x = y + 1 \\
S(\langle v, x, y, z \rangle, \langle v', x', y', z' \rangle) & \triangleq x ' = x \wedge y' = y \wedge \\
& z' + v' = x' \wedge v' \leq y' \\
R(x) & = x
\end{align*}

>>>>

All we need to do in order to use {\bf [CT]} in any context %environment and regardless of the complexity of the loops's body 
is generate constraints over-approximating %abstracting 
the environment and the loop's body. %: for termination, we need over-approximating abstractions, whereas for non-termination 
As the generation of such constraints is standard and covered by several other works \cite{DBLP:conf/pldi/GrebenshchikovLPR12,DBLP:conf/pldi/GulwaniSV08}, 
we will not provide the full algorithm, but rather illustrate it through the example in Figure~\ref{fig:environment-model}.
Our technique proves the termination of each loop separately and in program order, 
with the exception of nested loops, whose termination we encode in the same formula.
Following the same design, let us assume that we have already proven the termination of loop $L_1$, and we are now trying to prove that $L_2$ and $L_3$ terminate, which we encode
as the conjunction of constraints on the right side of the figure:
%% =======
%% %
%% As the generation of such constraints is standard and covered by several
%% other
%% works~\cite{DBLP:conf/pldi/GrebenshchikovLPR12,DBLP:conf/pldi/GulwaniSV08},
%% we will not provide the full algorithm, but rather illustrate it through the
%% example in Figure~\ref{fig:environment-model}.  Let us assume that we have
%% already proven the termination of loops $L_1$ and $L_3$, and we are now
%% trying to prove that $L_2$ terminates, which we encode as the conjunction of
%% constraints on the right-hand side of the figure:
%% %
%% >>>>>>> ee2c21bbc574ab36e7d6739c0e82c7a03d930789
\begin{itemize}
\item $W_1$ and $W_2$ are (over-approximating) inductive invariants for $L_1$ and $L_2$, respectively.
\item $S$ is a summary of $L_3$ that over-approximates the transitive closure of its transition relation.
\item If the formula is satisfiable, then $R_2$ and $R_3$ are ranking functions for $L_2$ and $L_3$, respectively.
\end{itemize}

Note that despite its intricacy, the constraint system is still a formula in the synthesis fragment.
Since the outer loop of this program does indeed terminate, the system is satisfiable, as witnessed by:
\begin{align*}
W_1(\langle v, x, y, z \rangle ) & \triangleq x \leq y + 1 \\
W_2(\langle v, x, y, z \rangle ) & \triangleq x = y + 1 \\
S(\langle v, x, y, z \rangle, \langle v', x', y', z' \rangle) & \triangleq x ' = x \wedge y' = y \wedge \\
& z' + v' = x' \wedge v' \leq y' \\
R_2(x) & = x\\
R_3 & = ..
\end{align*}
